---
title: "Public Safety Open Data Portal - Part I"
author: "Dov Chelst"
date: '2017-06-04'
output:
  blogdown::html_page:
    toc: yes


description: Background and Minor Critique
slug: public-safety-open-data-portal
tags:
- police
- Police Foundation
- White House
- ~
categories: []
topics: []
---

## Introduction & Critique

The [Police Data Initiative](https://www.policedatainitiative.org/datasets/)
website contains data supplied by 135 
[participating agencies](https://www.policedatainitiative.org/participating-agencies/) 
as of today's date. There is likely an agency near you that you might recognize.

In late 2015, the Tucson Police Department's Lieutenant Myron "Ron" Holubiak 
brought the Public Safety Open Data Portal to my attention. 
The White House's Police Data Initiative had begun in 
[May 2015](https://obamawhitehouse.archives.gov/blog/2015/05/18/launching-police-data-initiative) 
and Tucson PD had announced its participation at the 
[end of October](https://obamawhitehouse.archives.gov/blog/2015/10/27/police-data-initiative-5-month-update).

When I originally considered this topic, I hadn't planned to include a critique, 
yet when compared with other data portals, it has some flaws. 
Below are my impressions. 
I have not contacted the people managing the data portal for comment.  

- It's not _curated_. That means that the portal relies on each agency to 
control each dataset's quality. 
It doesn't look as if someone on the website reviews the data 
and provides feedback to agencies regarding their datasets' accuracy or 
completeness.  
- It's not (frequently) _updated_.
You may see an agency providing data on the portal, but providing more 
recent information on a city website. 
- It may not be _housed locally_. Generally, I would expect either a direct 
link to a file within the data portal. Alternatively, there might be descriptions 
of data stored elsewhere with direct links to those data files. 
In multiple instances, the Police Data Initiative simply links to a city's data
site and an interested researcher must hunt for the data there.

Here are some examples:

- **Boston Police Department** is listed as a participating agency, 
but shows no data on the site. Instead, there is a link to the 
[Legacy Boston Open Data Portal](https://data.cityofboston.gov/browse?tags=crime)
when you would rather go to the new site called 
[Analyze Boston](https://data.boston.gov/), where you can 
[download](https://data.boston.gov/dataset/6220d948-eae2-4e4b-8723-2dc8e67722a3/resource/12cb3883-56f5-47de-afa5-3b1cf61b257b/download/crime.csv) 
[crime incident reports from 2015 to present](https://data.boston.gov/dataset/crime-incident-reports-august-2015-to-date-source-new-system). 
These are _incidents_ from the department's records management system (RMS)
rather than _calls_ from the department's computer-aided dispatch (CAD) system.
It includes information on the time that an incident occurred but nothing 
about responding units or response times.  
On a positive note, incident locations contain latitudes and longitudes that 
appear quite accurate.

- **Atlanta Police Department** is also a participating agency. At the moment, 
the link is broken, but you can 
[download incident information here](http://www.atlantapd.org/i-want-to/crime-data-downloads). 
The same issue about "incidents versus calls"" applies to Atlanta. 
In fact, the call times appear intentionally truncated to the minute, 
with over 40% of records only showing the hour.  
- Within New Jersey, **Newark Police Department** references a link to a city
website that contains no police data, while the 
**Camden County Police Department**, which was an early participant in the 
initiative has no working link at all.

-----

## Atlanta 

```{r load_atlanta, echo=FALSE, message=FALSE}
# include some libraries
library(readxl)
library(readr) 
library(dplyr)
library(magrittr)
library(lubridate)
library(DT)
library(plotly)
# read in Excel data - download and unzip if necessary
atlanta.file <- "Atlanta-2017-FirstFourMonths.RData"
date.limits <- paste("2017", c(1, 5), 1, 0) %>% ymd_h
if(file.exists(atlanta.file)){
  load(atlanta.file)
} else{
  atlanta.page <- read_lines("http://www.atlantapd.org/i-want-to/crime-data-downloads")
  atlanta.link <- atlanta.page %>% grep("COBRA.*2017", ., value=TRUE)
  document.id <- gsub(".*id=(\\d+)\">COBRA-YTD2017.*", "\\1", atlanta.link)
  atlanta.url <- paste0("http://www.atlantapd.org/Home/ShowDocument?id=", 
                      document.id)
  download.file(atlanta.url, destfile="atlanta.zip", method="curl")
  unzip(zipfile="atlanta.zip")
  atlanta <- read_excel("COBRA-YTD2017.xlsx") %>%
    # convert text to dates and times and limit to 1/1/2017 through 4/30/2017
    mutate(occur_date = paste(occur_date, occur_time),
           poss_date = paste(poss_date, poss_time),
           rpt_date = mdy(rpt_date)) %>%
    mutate_at(vars(occur_date, poss_date), mdy_hms) %>%
    mutate(minute = minute(occur_date)) %>%
    mutate_at(vars(x, y, MaxOfnum_victims, beat), as.numeric) %>% 
    select(-matches("time")) %>%
    filter(between(occur_date, date.limits[1], date.limits[2]))
  save(atlanta, file="Atlanta-2017-FirstFourMonths.RData")
  file.remove("atlanta.zip")
  file.remove("COBRA-YTD2017.xlsx")
}

day.summary  <- atlanta %>% 
  count(occur_date) %$%
  summary(n)  
```

Below we take a look at incidents for Atlanta PD. 
First we load the information and limit to the first 4 months of 2017.
Next, we examine the recorded times for these 
`r formatC(nrow(atlanta), big.mark=",")` incidents. 
There were an average of `r round(day.summary[["Mean"]], 1)` incidents per day.
Focusing on the minutes shows that officers entering the information were
likely to round to the nearest hour, 30 minutes, or other increments, rather 
than entering an accurately recorded time.


```{r analyze_atlanta, echo=FALSE, message=FALSE}
atlanta.minutes <- atlanta %>%
  count(minute, sort=TRUE) %>%
  ungroup %>%
  dplyr::rename(count=n) %>%
  mutate(percent = count / sum(count),
         cumulative = cumsum(percent))
atlanta.minutes %>%
  datatable(rownames=FALSE) %>%
  formatPercentage(c('percent', 'cumulative'), 2) %>%
  formatCurrency('count', '', digits=0)

plot_ly(atlanta.minutes, x=~minute, y=~count, type='bar')

```

## Boston



```{r load_boston, echo=FALSE, message=FALSE}
# include some libraries
# read in Excel data - download and unzip if necessary
boston.file <- "Boston-2017-FirstFourMonths.RData"
if(file.exists(boston.file)){
  load(boston.file)
} else {
    boston.url <- "https://data.boston.gov/dataset/6220d948-eae2-4e4b-8723-2dc8e67722a3/resource/12cb3883-56f5-47de-afa5-3b1cf61b257b/download/crime.csv"
  boston <- read_csv(boston.url) %>% 
    filter(between(OCCURRED_ON_DATE, date.limits[1], date.limits[2])) %>%
    mutate(OFFENSE_CODE2=substr(OFFENSE_CODE, 1, 3)) %>%
    select(-YEAR, -MONTH, -DAY_OF_WEEK, HOUR)
  save(boston, file=boston.file)
 file.remove("boston-crime.csv")
}
day.summary2  <- boston %>% 
  count(date=date(OCCURRED_ON_DATE)) %$%
  summary(n)  
```

For simplicity, we apply the same limits to Boston PD's incidents that were
applied to Atlanta PD's incidents.
For Boston, we examine the recorded offense descriptions, focusing on UCR Part 
One crimes. 
There are a total of `r formatC(nrow(boston), big.mark=",")` incidents, and 
an average of `r round(day.summary2[["Mean"]], 1)` incidents per day.
There were 
`r formatC(sum(boston$UCR_PART=="Part One", na.rm=TRUE), big.mark=",")` 
Part One crimes.
Below is a quick count of crimes by group. _Note that the Boston PD appears to 
have removed all rape records from their public data set._

```{r boston_part1_crimes, echo=FALSE, message=FALSE}
boston %>% 
  filter(UCR_PART=="Part One") %>% 
  mutate(OFFENSE_CODE = gsub("^0+", "", OFFENSE_CODE),
         OFFENSE_CODE = substr(OFFENSE_CODE, 1, 1)) %>%
  count(OFFENSE_CODE, OFFENSE_CODE_GROUP) %>%
  ungroup %>%
  dplyr::rename(Count=n, Code=OFFENSE_CODE, Description=OFFENSE_CODE_GROUP) %>%
  datatable(options=list(rownames=FALSE, paging=FALSE)) %>%
  formatCurrency('Count', '', digits=0)

```

## Part II: 911 Calls New Orleans LA & Los Angeles CA (later this month)