---
title: "UCR 2015 Available (aka data set 36789)"
author: "Dov Chelst"
date: '2017-06-11'
description: ''
draft: yes
slug: ucr-2015-available-aka-data-set-36789
tags: []
categories:
- police
- crime
- UCR
topics: []
---


## Basic Information
The FBI released their 
[annual report](https://ucr.fbi.gov/crime-in-the-u.s/2015/crime-in-the-u.s.-2015) 
on Crime in the U.S. for 2015 
[last September](https://www.fbi.gov/news/pressrel/press-releases/fbi-releases-2015-crime-statistics).
It took a little longer than last year for an improved 
[data set (36789)](http://www.icpsr.umich.edu/icpsrweb/NACJD/studies/36789?searchSource=revise&q=36789) 
with to appear on the 
[NACJD website](http://www.icpsr.umich.edu/icpsrweb/NACJD/index.jsp). 

As I mentioned in a previous post, the main advantage of this data set is that 
it includes counts of _offenses cleared_ for each reporting agency along with
_offenses reported_. If you would like to have a conversation, 
there are reasons why this information may be inaccurate for most agencies in
[Washington state](http://www.waspc.org/statistics-reports). 
You might like this 
[article](http://www.news5cleveland.com/longform/cleveland-police-struggle-to-solve-murder-cases) 
by 
[Sarah Buduson](http://www.news5cleveland.com/about-us/staff/investigator-sarah-buduson-biography) at 
[News 5](http://www.news5cleveland.com/) 
on murder clearance rates in Cleveland with embedded video interviews. It also
includes comparisons with other cities of similar size such as New Orleans.

## Seven Cities

```{r load_libraries_data, echo=FALSE, message=FALSE}
library(tidyverse)
library(stringr)
library(magrittr)
library(DT)
data.dir <- "../../data"
load(file.path(data.dir, "UCR2015-BlogPost.RData"))

ucr.small %<>%
  select(-ori.code, -year, -months.reported, -agency.state.name, 
         -mailing.address.line.2) %>%
  dplyr::rename(city=agency.name, state=numeric.state.code) %>%
  mutate_at(vars(-matches("city|state")), as.numeric) %>%
  mutate(state=gsub(".*?\\s+(.*)", "\\1", state), 
         population = population.1 + population.2 + population.3) %>%
  select(-matches("[.]\\d$")) %>%  
  mutate_at(vars(city, state), str_to_title) 

crime.order <- c("murder.manslaughter", "rape", "robbery", 
                 "aggravated.assault", "burglary", "larceny", "vehicle.theft")

```

Along with describing the available data, I thought I'd use it a little bit.
Let's take a look at the UCR information associated with the four agencies
that I either discussed recently or plan to examine in the near future: 
Atlanta, Boston, Los Angeles, and New Orleans. 
In addition, if a basic consumer might not care about clearances, I'll include
three reasonably sized agencies whose data is not available on the FBI's 
website, but can be found in the new release: Ithaca (NY), Longmont (CA), 
and Schenectady (NY). 


### Crime Rates
Below are crime rates measured as crimes per 100,000.

```{r crime_rates, echo=FALSE, message=FALSE}
ucr.rates <- ucr.small %>%
  select(-matches("cleared")) %>%
    gather(crime, rate, rape:murder.manslaughter) %>%
  mutate(rate= round(10^5 * rate / population), 
         crime=factor(crime, levels=crime.order, ordered=TRUE)) %>%
  spread(crime, rate) %>%
  mutate(violent.crime=murder.manslaughter + rape + robbery +  aggravated.assault,
         property.crime = burglary + larceny + vehicle.theft) %>%
  select(city, state, population,
         violent.crime, murder.manslaughter:aggravated.assault,
         property.crime, burglary:vehicle.theft) %>%
  arrange(city)

names(ucr.rates) <- names(ucr.rates) %>% 
  gsub("[.]", " ", .) %>% 
  str_to_title

ucr.rates %>%
  datatable(rownames=FALSE, options=list(paging=FALSE)) %>%
  formatCurrency(-c(1:2), '', digits=0)

```

### Crimes and Clearances
Belw are counts of offenses and clearances by city and type of crime. 
Clearance percentages are calculated as the fraction of clearances relative to 
offenses.
The table allows you to limit observations to a single city or a single crime. 
You may note that New Orleans has the lowest murder clearance rate of the group.
Boston has the lowest robbery clearance rate of the group.

```{r crime_and_clearances, echo=FALSE, message=FALSE}
ucr.crimes <- ucr.small %>%
  gather(crime, count, rape:murder.manslaughter.cleared) %>%
  mutate(type=ifelse(grepl("clear", crime), "clearances", "offenses"),
         crime = gsub("[.]cleared", "", crime),
         crime = factor(crime, levels=crime.order, ordered=TRUE)) %>% 
  spread(type, count) %>%
  mutate(percentage = clearances / offenses) %>%
  select(city, state, population, crime, offenses, clearances, percentage) %>%
  arrange(city, state, crime) %>%
  mutate(crime = gsub("[.]", " ", crime),
         crime = str_to_title(crime))

ucr.crimes %>%
  set_names(str_to_title(names(ucr.crimes))) %>%
  datatable(rownames=FALSE, filter='top') %>%
  formatCurrency(c("Population", "Offenses", "Clearances"), '', digits=0) %>%
  formatPercentage('Percentage', digits=1)

