---
title: "Public Safety Open Data Portal - Part I"
author: "Dov Chelst"
date: '2017-05-23'
description: 'Background and Minor Critique'
draft: yes
publishdate: '2017-05-23'
slug: public-safety-open-data-portal
tags:
- police
- Police Foundation
- White House
- ~
categories: []
topics: []
---


The [Police Data Initiative](https://www.policedatainitiative.org/datasets/)
website contains data supplied by 135 
[participating agencies](https://www.policedatainitiative.org/participating-agencies/) 
as of today's date. There is likely an agency near you that you might recognize.

In late 2015, the Tucson Police Department's Lieutenant Myron "Ron" Holubiak 
brought the Public Safety Open Data Portal to my attention. 
The White House's Police Data Initiative had begun in May 2015 and 
Tucson PD had announced its participation at the end of October.

When I originally considered this topic, I hadn't planned to include a critique, 
yet when compared with other data portals, it has some flaws. 
Below are my impressions. 
I have not contacted the people at the data portal for comment.  

- It's not _curated_. That means the information is as good as what an agency 
wishes to provide. It doesn't look as if someone on the website reviews the data 
and provides feedback to an agency regarding their data's accuracy or 
completeness.  
- It's not (frequently) _updated_.
You may see an agency providing data on the portal, but 
providing more recent information on a city website. 
- It may not be _housed locally_. Generally, I would expect either a direct 
link to a file within the data portal. Alternatively, there might descriptions 
of data stored elsewhere with direct links to those data files. 
In multiple instances, the Police Data Initiative simply links to a city's data
site and an interested researched must search for the data there.

Here are some examples:

- **Boston Police Department** is listed as a participating agency, 
but shows no data on the site. 
Instead, there is a link to the 
[Legacy Boston Open Data Portal](https://data.cityofboston.gov/browse?tags=crime)
when you would rather go to the new site called 
[Analyze Boston](https://data.boston.gov/), where you can 
[download](https://data.boston.gov/dataset/6220d948-eae2-4e4b-8723-2dc8e67722a3/resource/12cb3883-56f5-47de-afa5-3b1cf61b257b/download/crime.csv) 
[crime incident reports from 2015 to present](https://data.boston.gov/dataset/crime-incident-reports-august-2015-to-date-source-new-system). 
These are _incidents_ from the department's records management system (RMS)
rather than _calls_ from the department's computer-aided dispatch (CAD) system.
It includes information on the time an incident occurred but nothing about 
responding units or response times.  
- **Atlanta Police Department** is also a participating agency. At the moment, 
the link is broken, but you can 
[download incident information here](http://www.atlantapd.org/i-want-to/crime-data-downloads). 
The same issue about "incidents versus calls"" applies to Atlanta. 
In fact, the call times appear intentionally truncated to the minute, 
with over 40% of records only showing the hour.  
- 


-----

Below we take a look at call for Atlanta PD. First we load the information.

```{r load_atlanta, echo=TRUE, message=FALSE}
# include some libraries
library(readxl)
library(readr) 
library(dplyr)
library(lubridate)
library(DT)
library(plotly)
# read in Excel data - download and unzip if necessary
incidents.file <- "COBRA-YTD2017.xlsx"
if(!file.exists(incidents.file)){
  atlanta.url <- "http://www.atlantapd.org/Home/ShowDocument?id=1150"
  download.file(atlanta.url, destfile="atlanta.zip", method="curl")
  unzip(zipfile="atlanta.zip")
}
# convert text to dates and times
atlanta <- read_excel(incidents.file) %>%
  mutate(occur_date = mdy(occur_date),
         occur_time = hms(occur_time),
         minute = minute(occur_time))
```

Next, we examine the recorded times for these 
`r formatC(nrow(atlanta), big.mark=",")` incidents. 
Focusing on the minutes shows that officers entering the information were
likely to round to the nearest hour, 30 minutes, or other increments, rather 
than entering an accurately recorded time.


```{r analyze_atlanta, echo=TRUE, message=FALSE}
atlanta.minutes <- atlanta %>%
  count(minute, sort=TRUE) %>%
  ungroup %>%
  dplyr::rename(count=n) %>%
  mutate(percent = count / sum(count),
         cumulative = cumsum(percent))
atlanta.minutes %>%
  datatable(rownames=FALSE) %>%
  formatPercentage(c('percent', 'cumulative'), 2) %>%
  formatCurrency('count', '', digits=0)

plot_ly(atlanta.minutes, x=~minute, y=~count, type='bar')

```

```{r clean_up, echo=TRUE}


```