--- 
title: 'NFIRS 2015 is Vastly Improved: Just Ask FEMA' 
author: "Dov Chelst" 
date: '2018-02-11'
description: ''
slug: nfirs-2015-improved
tags:
- nfirs
- fire
categories: []
topics: []
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE) 
load("NFIRS-2015-BasicAnalysis.RData")
library(DT)
library(tidyverse)
library(magrittr)

total.calls <- incident.by.type %$% sum(nn)
SubType <- function(type1){
  sum1 <- incident.by.type %>% 
    filter(INC_TYPE %in% type1$INC_TYPE) %$% 
    sum(nn)
  return(sum1)}
RoundThousand <- function(number1){
  rounded.number <- number1 %>% 
    round(-3) %>% 
    formatC(big.mark=",", format="d")
  return(rounded.number)
}
fire.calls <- SubType(fire.types)
medical.calls <- SubType(medical.types)
```

## Introduction

Data from the National Fire Incident Reporting System (NFIRS) is now available
from the U.S. Fire Administration (USFA) for 2015. The agency collects data
reported from the majority of fire departments across the country. While working
with fire departments, in my current role, I've observed that most reporting
agencies provide detailed information about every incident along with times
associated with each of the incident's responding units. Additional details
accompany serious fires, mutual aid responses, casualties (injuries and deaths),
and assorted special circumstances (such as hazardous materials and wildfires).

In the past, the USFA's public data release (PDR) would only contain fire
incidents, which amounts to a small but significant fraction (8.7% for 2015) of
all incidents involving fire departments. Now, you can request information about
__all__ incidents; however, unit-level information still remains unavailable.
Last May, I received a full data set for 2015 and a matching set of data for
2014 as well. You can request your own data set
[here](https://www.usfa.fema.gov/data/statistics/order_download_data.html).

I'm excited about the update as this new expanded data set places NFIRS on par
with other similar national data sets such as the National Highway and Traffic
Safety Administration's ([NHTSA](https://www.nhtsa.gov/)) National Emergency
Medical Services Information System ([NEMSIS](https://nemsis.org/)) that records
emergency medical incidents involving public or private ambulance services
throughout the country. Moreover, it allows a researcher to get a fuller picture
of any single agency as well as the activities of the fire service as a whole.
2015's data includes approximately 24.8 million records, while 2014's original
"fires only" data release contained approximately 2.1 million records. Focusing
on the country's largest fire department, the New York Fire Department (NYFD),
NFIRS 2015 data includes approximately 557,000 incidents, but only 80,000
fire-specific incidents.

I'd like to take a few moments to discuss the new data set. NFIRS 2015 data 
presents challenges that arise whenever one deals with larger data sets. 
In addition, I recently noticed that many agency's data submissions are incomplete.
Nevertheless, I'd like to provide some interesting summary information. Some of 
this will leverage the newly included information, highlighting medical 
incidents. Another will emphasize casualty data which is an important focus of 
the USFA's statistical analysis; but I'll expand the scope to include fire 
injuries as well as fire deaths. At the end, I will include some foods for 
thought, links to my own analysis for the technically savvy, and hints at 
further analysis to come.

## Difficulties Associated With Data Size

I usually conduct the majority of my data analysis using the [R statistical
programming language](https://www.r-project.org/) on a single workstation. This
works fine for smaller data sets that can be loaded comfortably into a
computer's RAM. However, the NFIRS 2015 data release includes more than 10 files
and contains more than 7 Gigabytes of information. Even looking at files
individually, the primary data file alone exceeds 4 Gigabytes in size. I'd like
to discuss some reasonable approaches to this specific problem that I plan to
explore going forward along with others that I've already discarded.

- __Ignoring the problem__: I'm certainly not loading all the data into memory
and storing the information as one RData file. It's just too big for my system.
- __Standard Relational Database__: It seems clear that USFA is using a standard
database to manage it's own NFIRS data. Here's why I won't be reproducing that
method.
    - I've explored using MySQL and/or MariaDB for this purpose. While useful
    for me in general, the data loading process for files of this size is still
    a bit involved.
    - For a database of this size, with large individual tables, running queries
    on my home computer is still quite slow.
    - Cloud hosting is possible but adds an extra layer of effort with limited
    benefit as I'm not sharing the data with many users.
    
- __Limiting Data__: To accomplish the analysis incorporated in later sections,
I've chosen to simply read in a _very limited_ portion of the data set. Careful
decisions about what is absolutely necessary for an analysis can alleviate
problems with data size. At this point, I'm just using *read_csv* from the
[readr](http://readr.tidyverse.org/) package. It allows me to selectively load
11 out of 40 columns from the NFIRS 2015 basic incident table. The resulting RDS
file, which is a standard internal R format for a single data set (or "data
frame") is less than 90 Megabytes when compared to 4 Gigabytes. Nevertheless,
the file still took over 4 minutes to load. Another alternative is using
*read.csv.sql* from the
[sqldf](https://github.com/ggrothendieck/sqldf/blob/master/README.md#Overview)
package.
- __Drill__: For the next NFIRS-related post, I will likely use [Apache
Drill](https://drill.apache.org/) for basic queries. It can also convert the
current text-delimited data files into a more efficient columnar 
[Parquet](https://parquet.apache.org/) format. I've been experimenting with
Drill recently and it's worked quite well for me.  
- __Athena__: I'm a current Amazon Web Services user and 
[Athena](https://aws.amazon.com/athena/) would allow me a cost-effective way to 
leverage cloud computing and my [Simple Storage Service
(S3)](https://aws.amazon.com/s3/) to perform a complicated analysis in a
reasonable amount of time. I haven't explored this option adequately to offer an
informed opinion on it.

## Basic Information

The 2015 NFIRS data set includes approximately `r RoundThousand(total.calls)`
incidents. 

In the past, the public data set was limited to only `r nrow(fire.types)` fire
incident types. In 2015, these fire incident types represented approximately 
`r RoundThousand(fire.calls)` calls. The following table shows these types 
sorted from most to least frequent.
```{r fire_types, echo=FALSE, message=FALSE}
incident.by.type %>%
  select(-percent) %>%
  filter(INC_TYPE %in% fire.types$INC_TYPE) %>%
  arrange(desc(nn)) %>%
  datatable(rownames=FALSE, 
            colnames=c("Incident Type", "Description", "Frequency")) %>%
  formatRound('nn', 0)
```

It's reasonable to expect that the departments with the largest number of calls
also manage areas with the largest populations. While there are 
`r nrow(call.by.agency) %>% formatC(big.mark=",", format="d")` agencies reporting
at least some of their calls to the national system, below is a list of the 100
agencies with the most reported calls. While there are 50 states in the U.S.,
these 100 agencies are concentrated in 31 states. For example, no agencies are
included from Michigan or New Jersey.

```{r calls_by_agency, echo=FALSE, message=FALSE}
call.by.agency %>%
  arrange(desc(nn)) %>%
  head(100) %>%
  datatable(rownames=FALSE, 
            colnames=c("State", "Fire Department ID", "Name", "Total Calls")) %>%
  formatRound('nn', 0)
```

## Medical Percentage

Without examining populations in detail, it should be clear that the number of
calls is not always proportional to an area's population. For example,
Fayetteville Fire and EMS reported a similar number of calls as the San
Francisco Fire Department. However, San Francisco's population is approximately
865,000 people while Fayetteville (NC)'s population is approximately 204,000
people. The main difference is that Fayetteville's fire department reported more
medical calls to NFIRS than San Francisco's Fire Department.[^1] A way to compare
reporting by each agency is to convert separate calls into fire calls, medical
calls, and the rest. Then, we calculate the percentage of reported calls that
were fire calls (as per prior year's releases) or medical calls.

Overall, there were `r RoundThousand(medical.calls)` medical calls which was
`r round(100 * medical.calls / total.calls, 1)`% of total calls. The next table
shows calls and associated percentages for all agencies that reported at least 
1,000 calls in 2015.[^2]

```{r medical_proportion, echo=FALSE, message=FALSE}
incident.type.by.agency %>%
  mutate(incident.type2 = case_when(
    INC_TYPE %in% medical.types$INC_TYPE ~ "medical",
    INC_TYPE %in% fire.types$INC_TYPE ~ "fire", 
    TRUE ~ "other")) %>%
  count(STATE, FDID, FD_NAME, incident.type2, wt=n) %>%
  spread(incident.type2, nn) %>%
  mutate(total=medical + fire + other, 
         fire.pct = fire / total, 
         medical.pct = medical / total) %>% 
  arrange(desc(total)) %>%
  filter(total >= 1000) %>%
  select(-other) %>%
  datatable(rownames=FALSE, 
            colnames=c("State", "FDID", "Name", 
                       "Fire", "Medical",   #"Other Calls", 
                       "Total",
                       "Fire %", "Medical %"),
            filter="top") %>%
  formatRound(c('fire', 'medical', # 'other',
                'total'), 0) %>%
  formatPercentage(c("fire.pct", "medical.pct"))
```

### Food for Thought

In upcoming posts, we'll examine:  

- Problems associated with the NFIRS 2015 data set  
- Components of NFIRS associated with casualties (death and injury)  
- Merging NFIRS data with census information to determine associated rates  

[^1]: Keep in mind that counts of "reported calls" may differ from actual calls. 
This will be explored in a future post.
[^2]: The table's fire department names relies on a table supplied by NFIRS. The
list is just to correct the spelling of each entry manually before publishing
this post.